---
title: "PCA & MDS Visualization of Suicidality Embeddings"
author: "Zhuojun"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: html_document
params:
  dataset_choice: "d2"
---

## Overview

This R Markdown document reproduces the exploratory visualization pipeline that was originally written as a plain R script.  It

1. Loads sentence‐embedding matrices and model predictions;
2. Runs a PCA on the embeddings and keeps enough components to explain ≥85 % of the variance;
3. Calculates a two‑dimensional MDS from the PCA space;
4. Generates a set of hex‑bin and scatter plots that summarize relationships between the low‑dimensional representation and suicidality ratings (expert and model‐predicted) as well as observed PHQ‑9 totals.

> **Prerequisites**  
> • The files `suidat/textpack_suicide_emb.rds` and `suidat/textpack_suicide_pred.rds` must exist relative to the working directory.  
> • The YAML parameter `dataset_choice` selects the subset (`"d1"`, `"d2"`, or `"d3"`).  Edit it at the top of this file or supply `rmarkdown::render("pca_mds_visualization.Rmd", params = list(dataset_choice = "d3"))`.

## Packages

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(hexbin)
library(magrittr)
library(dplyr)
library(tibble)
library(tidymodels)
library(ggplot2)
library(stats)       # dist(), cmdscale()
library(ggthemes)
library(patchwork)
library(tidyr)
```

## Core function

```{r functions}
run_pca_mds_visualization <- function(dataset_choice = "d2") {
  # ---- Load & merge data ----------------------------------------------------
  embdf <- readRDS('suidat/textpack_suicide_emb.rds')$texts$texts
  df    <- readRDS('suidat/textpack_suicide_pred.rds')
  df    <- dplyr::bind_cols(df, embdf)

  # ---- Filter ---------------------------------------------------------------
  df <- df[df$dataset == dataset_choice, ]

  # Keep metadata & embeddings
  df <- df[, c(3, 4, 7, 8:1031)]
  names(df)[2:3] <- c('expert_suicidality', 'LBA_suicidality')

  # ---- PCA (≥85 % variance) --------------------------------------------------
  pca_recipe <- recipe(~ ., data = df[, 4:1027]) %>%
    step_pca(all_predictors(), threshold = 0.85) %>%
    prep()
  df_pca <- bake(pca_recipe, new_data = NULL)

  # ---- MDS on Euclidean distances in PCA space ------------------------------
  dissimilarity <- dist(as.matrix(df_pca))
  mds_result    <- cmdscale(dissimilarity, k = 2)
  df_mds <- as_tibble(mds_result, .name_repair = "unique") %>%
    rename(MDS1 = ...1, MDS2 = ...2)

  # ---- Combine & standardise ------------------------------------------------
  df_combined <- bind_cols(
    df_pca,
    df_mds,
    df$PHQtot,
    df$expert_suicidality,
    df$LBA_suicidality
  ) %>%
    mutate(across(everything(), ~ as.numeric(scale(.x))))
  ncoltemp <- ncol(df_pca) + ncol(df_mds) + 1
  names(df_combined)[ncoltemp:ncol(df_combined)] <- c('PHQtot',
                                                      'expert_suicidality',
                                                      'LBA_suicidality')

  # Optional: write a tidy CSV for Python/Seaborn users
  write.csv(df_combined, paste0('plotseaborn_', dataset_choice, '.csv'),
            row.names = FALSE)

  # ---- Visualisation --------------------------------------------------------
  common_limits <- c(-3, 3)
  plot_list <- list(
    MDSExpertRat = ggplot(df_combined, aes(MDS1, MDS2)) +
      stat_summary_hex(aes(z = expert_suicidality), bins = 10, fun = mean) +
      scale_fill_gradient2(low = "blue", mid = "lightgrey", high = "red",
                           name = "Avg Expert Suicidality", midpoint = 0,
                           limits = common_limits) +
      labs(title = sprintf("Hexbin: Expert Suicidality (%s)", dataset_choice)) +
      theme_minimal(),

    PCAExpertRat = ggplot(df_combined, aes(PC1, PC2)) +
      stat_summary_hex(aes(z = expert_suicidality), bins = 10, fun = mean) +
      scale_fill_gradient2(low = "blue", mid = "lightgrey", high = "red",
                           name = "Avg Expert Suicidality", midpoint = 0,
                           limits = common_limits) +
      labs(title = sprintf("Hexbin: Expert Suicidality (PCA – %s)",
                           dataset_choice)) +
      theme_minimal(),

    MDSLBA = ggplot(df_combined, aes(MDS1, MDS2)) +
      stat_summary_hex(aes(z = LBA_suicidality), bins = 10, fun = mean) +
      scale_fill_gradient2(low = "blue", mid = "lightgrey", high = "red",
                           name = "Avg LBA Suicidality", midpoint = 0,
                           limits = common_limits) +
      labs(title = sprintf("Hexbin: LBA Suicidality (%s)", dataset_choice)) +
      theme_minimal(),

    PCALBA = ggplot(df_combined, aes(PC1, PC2)) +
      stat_summary_hex(aes(z = LBA_suicidality), bins = 10, fun = mean) +
      scale_fill_gradient2(low = "blue", mid = "lightgrey", high = "red",
                           name = "Avg LBA Suicidality", midpoint = 0,
                           limits = common_limits) +
      labs(title = sprintf("Hexbin: LBA Suicidality (PCA – %s)",
                           dataset_choice)) +
      theme_minimal(),

    ScatterExpertLBA = ggplot(df_combined, aes(MDS1, MDS2,
                                               colour = LBA_suicidality,
                                               size   = expert_suicidality)) +
      geom_point(alpha = 0.6) +
      scale_colour_gradient2(low = "blue", mid = "lightgrey", high = "red",
                             midpoint = 0, limits = common_limits) +
      labs(title = sprintf("MDS Scatter (%s)", dataset_choice),
           colour = "LBA Suicidality",
           size   = "Expert Suicidality") +
      theme_minimal()
  )

  return(plot_list)
}
```

## Generate plots for the selected dataset

```{r run-analysis, results='hide'}
plots <- run_pca_mds_visualization(params$dataset_choice)
```

### Scatter plot (Expert vs LBA) – *`r params$dataset_choice`*

```{r scatter-expert-lba, fig.width=7, fig.height=5}
plots$ScatterExpertLBA
```

### Hex‐bin comparison (Expert vs LBA) – *`r params$dataset_choice`*

```{r hexbin-expert-lba, fig.width=12, fig.height=5}
plots$MDSExpertRat + plots$MDSLBA
```


## Compare multiple datasets

The chunk below reproduces the composite figure from the original script:

```{r multi-dataset d1, eval=TRUE, fig.width=12, fig.height=5}
out1 <- run_pca_mds_visualization("d1")
out2 <- run_pca_mds_visualization("d2")
out3 <- run_pca_mds_visualization("d3")

out1$MDSLBA + out1$PCALBA
```

```{r multi-dataset d2, eval=TRUE, fig.width=12, fig.height=5}
out2$MDSLBA + out2$PCALBA

```

```{r multi-dataset d3, eval=TRUE, fig.width=12, fig.height=5}

 (out3$MDSLBA + out3$PCALBA)

```
